Loop-Aware Optimizations in PyPy's Tracing JIT

### 2011-06-20 talk/iwtc11/paper.tex
Add benchmark table - too big as of now
squeeze a bit
intending to make benchmarks more reasonable (will update the code as well)

\usepackage{ifthen}
\usepackage{fancyvrb}
\usepackage{color}
\usepackage{ulem}
\usepackage{xspace}
\usepackage{epsfig}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage[utf8]{inputenc}
\usepackage{setspace}

@@

The loop peeling optimization was implemented in the PyPy
framework. That means that the JIT-compilers generated for all
interpreters implemented within PyPy now can take advantage of
it. Benchmarks have been executed for a few different interpreters and
we see improvements in several cases. The ideal loop for this optimization
would be short numerical calculations with no failing guards and no
external calls. Larger loops involving many operations on complex objects
typically benefit less from it. Loop peeling never makes performance worse, in
the worst case the peeled loop is exactly the same as the preamble. Therefore we
chose to present benchmarks of small numeric kernels where loop peeling can show
its use.

@@

\subsection{Python}
The Python interpreter of the PyPy framework is a complete Python
version 2.7 compatible interpreter. A set of numerical
calculations were implemented in both Python and in C and their
runtimes compared. The benchmarks are
\begin{itemize}
\item {\bf sqrt}: approximates the square root of $y$ as $x_\infty$
  with $x_0=y/2$ and $x_k = \left( x_{k-1} + y/x_{k-1} \right) /
  2$. There are three different versions of this benchmark where $x_k$
  is represented with different type of objects: int's, float's and
  Fix16's. The latter, Fix16, is a custom class that implements
  fixpoint arithmetic with 16 bits precision. In Python there is only
  a single implementation of the benchmark that gets specialized
  depending on the class of it's input argument, $y$, while in C,
  there are three different implementations.
\item {\bf conv3}: one-dimensional convolution with fixed kernel-size $3$.
\item {\bf conv5}: one-dimensional convolution with fixed kernel-size $5$.
\item {\bf conv3x3}: two-dimensional convolution with kernel of fixed
  size $3 \times 3$ using a custom class to represent two-dimensional
  arrays.
\item {\bf dilate3x3}: two-dimensional dilation with kernel of fixed
  size $3 \times 3$. This is similar to convolution but instead of
  summing over the elements, the maximum is taken. That places a
  external call to a max function within the loop that prevents some
  of the optimizations.
\item {\bf sobel}: a low-level video processing algorithm used to
  locate edges in an image. It calculates the gradient magnitude
  using sobel derivatives. In Python the algorithm is implemented
  on top of a custom image class that is specially designed for the
  problem. It ensures that there will be no failing guards, and makes
  a lot of the two dimension index calculations loop invariant. The
  intention here is twofold. It shows that the performance-impact of
  having wrapper classes giving objects some application-specific
  properties can be negligible. This is due to the inlining performed
  during the tracing and the allocation removal of the index objects
  introduced. It also shows that it is possible to do some low-level
  hand optimizations of the Python code and hide those optimization
  under a nice interface without loosing performance.
\end{itemize}

@@

We can observe that PyPy (even without loop peeling) is orders of magnitude
faster than either CPython or Psyco. This is due to the JIT compilation
advantages and optimizations we discussed in XXX [ref to other paper]. Loop
peeling gives an additional XXX on average, which makes benchmark times
comparable with native-compiled C code. Missing performance we attribute to
the relative immaturity of PyPy's JIT assembler backend as well as missing
optimizations, like instruction scheduling.

>>>>

\usepackage{ifthen}
\usepackage{fancyvrb}
\usepackage{color}
\usepackage{ulem}
\usepackage{xspace}
\usepackage{epsfig}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage[utf8]{inputenc}
\usepackage{setspace}
\usepackage{relsize}

@@

The loop peeling optimization was implemented in the PyPy
framework. That means that the JIT-compilers generated for all
interpreters implemented within PyPy now can take advantage of
it. Benchmarks have been executed for a few different interpreters and
we see improvements in several cases. The ideal loop for this optimization
would be short numerical calculations with no failing guards and no
external calls. Larger loops involving many operations on complex objects
typically benefit less from it. Loop peeling never makes runtime performance worse, in
the worst case the peeled loop is exactly the same as the preamble. Therefore we
chose to present benchmarks of small numeric kernels where loop peeling can show
its use.

@@

\subsection{Python}
The Python interpreter of the PyPy framework is a complete Python
version 2.7 compatible interpreter. A set of numerical
calculations were implemented in both Python and in C and their
runtimes compared. The benchmarks are
\begin{itemize}
\item {\bf sqrt}: approximates the square root of $y$ as $x_\infty$
  with $x_0=y/2$ and $x_k = \left( x_{k-1} + y/x_{k-1} \right) /
  2$. There are three different versions of this benchmark where $x_k$
  is represented with different type of objects: int's, float's and
  Fix16's. The latter, Fix16, is a custom class that implements
  fixpoint arithmetic with 16 bits precision. In Python there is only
  a single implementation of the benchmark that gets specialized
  depending on the class of it's input argument, $y$, while in C,
  there are three different implementations.
\item {\bf conv3}: one-dimensional convolution with fixed kernel-size $3$.
\item {\bf conv5}: one-dimensional convolution with fixed kernel-size $5$.
\item {\bf conv3x3}: two-dimensional convolution with kernel of fixed
  size $3 \times 3$ using a custom class to represent two-dimensional
  arrays.
\item {\bf dilate3x3}: two-dimensional dilation with kernel of fixed
  size $3 \times 3$. This is similar to convolution but instead of
  summing over the elements, the maximum is taken. That places a
  external call to a max function within the loop that prevents some
  of the optimizations.
\item {\bf sobel}: a low-level video processing algorithm used to
  locate edges in an image. It calculates the gradient magnitude
  using sobel derivatives. 
\end{itemize}

The sobel and conv3x3 benchmarks are implemented
on top of two different two-dimensional array classes: Array2D
and NoBorderPadded. Array2D is 
a simple straight forward implementation providing 2 dimensionall
indexing with out of bounds checks. NoBorderPadded is optimized for
this specific application.
It ensures that there will be no failing guards, and makes
a lot of the two dimension index calculations loop invariant. The
intention here is twofold. It shows that the performance-impact of
having wrapper classes giving objects some application-specific
properties can be negligible. This is due to the inlining performed
during the tracing and the allocation removal of the index objects
introduced. It also shows that it is possible to do some low-level
hand optimizations of the Python code and hide those optimization
under a nice interface without loosing performance.

@@

We can observe that PyPy (even without loop peeling) is orders of magnitude
faster than either CPython or Psyco. This is due to the JIT compilation
advantages and optimizations we discussed in XXX [ref to other paper]. Loop
peeling gives an additional XXX on average, which makes benchmark times
comparable with native-compiled C code. Missing performance we attribute to
the relative immaturity of PyPy's JIT assembler backend as well as missing
optimizations, like instruction scheduling. Results:

\begin{figure}
\begin{center}
{\smaller
\begin{tabular}{|l|r|r|r|r|r|r|}
\hline
 & CPython & Psyco & PyPy no LP & PyPy & GCC -O2 & GCC -O3\\
\hline
NoBorder & 537.31 & 329.98 & 2.22 +- 0.03 & 2.17 +- 0.02 & - & -\\
\hline
NoBorder(iter) & 548.32 & 304.13 & 1.45 +- 0.03 & 1.47 +- 0.02 & - & -\\
\hline
NoBorder(range) & 534.64 & 317.34 & 1.34 +- 0.03 & 1.40 +- 0.04 & - & -\\
\hline
NoBorderPadded & 543.73 & 333.20 & 2.09 +- 0.12 & 1.93 +- 0.05 & - & -\\
\hline
NoBorderPadded(iter) & 546.70 & 309.32 & 1.21 +- 0.02 & 0.49 +- 0.02 & - & -\\
\hline
NoBorderPadded(range) & 550.92 & 318.33 & 1.12 +- 0.03 & 0.48 +- 0.01 & - & -\\
\hline
conv3(1e5) & 77.89 & 9.52 & 1.77 +- 0.06 & 0.68 +- 0.02 & 0.70 +- 0.05 & 0.59 +- 0.01\\
\hline
conv3(1e6) & 77.15 & 9.58 & 1.69 +- 0.01 & 0.77 +- 0.01 & 0.84 +- 0.05 & 0.74 +- 0.01\\
\hline
conv3x3(1000) & 23.72 & 12.77 & 0.07 +- 0.00 & 0.05 +- 0.03 & 0.24 +- 0.00 & 0.25 +- 0.01\\
\hline
conv3x3(3) & 23.85 & 12.77 & 0.10 +- 0.00 & 0.07 +- 0.00 & 0.27 +- 0.01 & 0.27 +- 0.01\\
\hline
conv5(1e5) & 122.54 & 16.67 & 1.86 +- 0.02 & 1.05 +- 0.03 & 1.03 +- 0.05 & 0.65 +- 0.01\\
\hline
conv5(1e6) & 125.77 & 16.80 & 1.92 +- 0.03 & 1.09 +- 0.02 & 1.07 +- 0.05 & 0.80 +- 0.01\\
\hline
dilate3x3(1000) & 23.29 & 12.99 & 0.41 +- 0.04 & 0.39 +- 0.01 & 0.26 +- 0.00 & 0.26 +- 0.01\\
\hline
sobel(NoBorderPadded) & 461.14 & 258.02 & 1.01 +- 0.03 & 0.48 +- 0.03 & - & -\\
\hline
sobel\_magnitude & - & - & - & - & 0.19 +- 0.01 & 0.20 +- 0.01\\
\hline
sobel\_uint8(NoBorderPadded) & 476.72 & 275.50 & 1.05 +- 0.01 & 0.51 +- 0.00 & - & -\\
\hline
sqrt(Fix16) & 744.35 & 421.65 & 3.93 +- 0.11 & 2.14 +- 0.02 & 0.97 +- 0.02 & 0.96 +- 0.01\\
\hline
sqrt(float) & 24.21 & 5.52 & 1.36 +- 0.03 & 1.00 +- 0.00 & 0.98 +- 0.00 & 0.98 +- 0.00\\
\hline
sqrt(int) & 20.84 & 1.78 & 2.26 +- 0.01 & 1.82 +- 0.01 & 0.81 +- 0.01 & 0.80 +- 0.01\\
\hline

\end{tabular}
}
\end{center}
\end{figure}
